---
title: "prospectr_soil_spectra"
author: "Jacob Nesslage"
date: "2024-03-06"
output: html_document
---
```{r}
require(prospectr)
```

Read in data 

```{r}
fp <- "Z:/Projects/Grayson_NB/Outputs/soil_spectra/soil_spectra_2022_withenvs.csv"
spec.l <- read_csv(fp)
spec.l
```

Wrangle data to wide format

```{r}
spec.wide <- spec.l %>% select (Wavelength, mean, Class1, Class2, Kit_Number, X, Y, transect,Moisture,Salinity,C,N,bulk_density) %>% 
  pivot_wider(., names_from = Wavelength, values_from = mean)
spec.wide
```

Isolate spectra and convert to matrix format

```{r convert spectra to matrix}
spc <- as.matrix(spec.wide[,12:2162])/100

# Plot the first spectrum
plot(x = as.numeric(colnames(spc)),
     y = spc[12, ],
     type = "l",
     lwd = 1.5,
     xlab = "Wavelength", 
     ylab = "Reflectance") 
```

Apply Savitzky-Golay smoothing and calculate derivatives of spectra

```{r Savitzky-Golay filtering and derivatives}
sg <- savitzkyGolay(X = spc, p = 3, w = 25, m = 0)

plot(x = as.numeric(colnames(sg)),
     y = sg[1, ],
     type = "l",
     lwd = 1.5,col="red",
     xlab = "Wavelength", 
     ylab = "Reflectance") 

sg_1st <- savitzkyGolay(X = spc, p = 3, w = 25, m = 1)

plot(x = as.numeric(colnames(sg_1st)),
     y = sg_1st[2, ],
     type = "l",
     lwd = 1.5,col="red",
     xlab = "Wavelength", 
     ylab = "Derivative of Reflectance") 



```

Apply conntinuum removal algorithm to spectra 

```{r continnum removal}
# type of data: 'R' for reflectance (default), 'A' for absorbance
cr <- continuumRemoval(X = sg, type = "R")
# plot of the 10 first abs spectra
matplot(as.numeric(colnames(sg)),
        t(sg[1:4,]),
        type = "l",
        lty = 1,
        ylim = c(0,1),
        xlab="Wavelength", 
        ylab="Reflectance")
matlines(as.numeric(colnames(sg)), lty = 1, t(cr[1:4, ]))
grid()
```

Principal Component Analysis of Spectra

```{r}
spec.pca <- prcomp(sg, scale = T)
spec.pca
pca_df <- spec.pca$rotation
```



Merge metadata with Savitzky-Golay spectra, derivative spectra, and continuum removal spectra

```{r merge dataframes together}
metadata <- spec.wide[,1:11]

final_sg <- cbind(metadata,sg) %>%
  filter(C != is.na(C))
write.csv(final_sg,"Z:/Projects/Grayson_NB/Outputs/soil_spectra/Grayson_savitzky_golay_spectra.csv")

final_sg_1st <- cbind(metadata,sg_1st) %>%
  filter(C != is.na(C))
write.csv(final_sg_1st,"Z:/Projects/Grayson_NB/Outputs/soil_spectra/Grayson_savitzky_golay_derivative.csv")

final_cr <- cbind(metadata,cr) %>%
  filter(C != is.na(C))
write.csv(final_cr,"Z:/Projects/Grayson_NB/Outputs/soil_spectra/Grayson_continuum_removal.csv")

final_pca <- cbind(metadata,t(pca_df)) %>%
  filter(C != is.na(C))
write.csv(final_pca,"Z:/Projects/Grayson_NB/Outputs/soil_spectra/Grayson_PCA.csv")
```

Calculate Pearson's Correlation for %C and each band for the three datasets

```{r}
# Calculate Pearson's correlation coefficient for each wavelength
correlations <- apply(as.data.frame(final_sg[,12:2138]), 2, function(column) cor(final_sg$C, column,method="pearson"))

# Create a data frame with wavelength and correlation values
correlation_data <- data.frame(Wavelength = names(correlations), Correlation = correlations)

# Summarize the results
summary(correlation_data)

#
plot(x = correlation_data$Wavelength,
     y = correlation_data$Correlation,
     type = "l",
     lwd = 1.5,col="red",
     xlab = "Wavelength", 
     ylab = "Correlation of Spectra and %C") 
```

```{r}
# Calculate Pearson's correlation coefficient for each wavelength
correlations <- apply(as.data.frame(final_sg_1st[,12:2138]), 2, function(column) cor(final_sg$C, column,method="pearson"))

# Create a data frame with wavelength and correlation values
correlation_data <- data.frame(Wavelength = names(correlations), Correlation = correlations)

# Summarize the results
summary(correlation_data)

#
plot(x = correlation_data$Wavelength,
     y = correlation_data$Correlation,
     type = "l",
     lwd = 1.5,col="red",
     xlab = "Wavelength", 
     ylab = "Correlation of 1st Derivative and %C") 
```

```{r}
# Calculate Pearson's correlation coefficient for each wavelength
correlations <- apply(as.data.frame(final_cr[,12:2138]), 2, function(column) cor(final_cr$C, column,method="pearson"))

# Create a data frame with wavelength and correlation values
correlation_data <- data.frame(Wavelength = names(correlations), Correlation = correlations)

# Summarize the results
summary(correlation_data)

#
plot(x = correlation_data$Wavelength,
     y = correlation_data$Correlation,
     type = "l",
     lwd = 1.5,col="red",
     xlab = "Wavelength", 
     ylab = "Correlation between Continuum Removal and %C")
```

```{r}
# Calculate Pearson's correlation coefficient for each wavelength
correlations <- apply(as.data.frame(final_pca[,12:2138]), 2, function(column) cor(final_pca$C, column,method="pearson"))

# Create a data frame with wavelength and correlation values
correlation_data <- data.frame(Wavelength = names(correlations), Correlation = correlations)

# Summarize the results
summary(correlation_data)

#
plot(x = correlation_data$Wavelength,
     y = correlation_data$Correlation,
     type = "l",
     lwd = 1.5,col="red",
     xlab = "Wavelength", 
     ylab = "Correlation between PCA and %C")
```

